Density-dependent selection and the limits of relative fitness

### 2018-04-04 manuscript.tex
l

For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, the Malthusian parameters $r_i$ will eventually start to decline due to crowding (e.g. \citealt[pp. 203]{begon_1990}). When population growth is density-regulated, selection can be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations. Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. Indeed, selection in crowded conditions is intimately connected to equilibrium density in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise, in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

Nevertheless, studies of density-dependent selection rarely incorporate reproductive excess. This requires that we model a finite, density-dependent excess, which is significantly more complicated zero (e.g. logistic) or infinite excess (e.g. Wright-Fisher). Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each type. In models with detailed age structure, the role of density is usually absorbed into one ``critical age-group'' that mediates the population's response to crowding. The existence of a reproductive excess implies that the critical age-group would have to occur at a pre-reproductive age. As a result, the focus shifts to juvenile density, obscuring the crucial connection between pre-reproductive-age selection, and post-reproductive-age density. For example, the result that selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994} is a form of the classical $K$-selection ideas discussed above applied to the critical age-group, and says nothing about other age-groups.

We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$ (and infinite reproductive excess). Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery.

We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.

@@

We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce. All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$. We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations around these two expectations.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these propagules secure the territories without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, and at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness). The third term, $A_i c_i/\overline{c}$, represents competitive victories in territories where two or more $i$ type propagules are present. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Two types are present with $c_1=1$, $c_2=1.5$ and $l_2/l_1=0.1$. Simulations are conducted as follows: $x_1,x_2$ values are sampled $U=10^5$ times from Poisson distributions with respective means $l_1,l_2$, and the victorious type in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/(c_1 x_1 + c_2 x_2)$.  Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the failure of the classic lottery model at low density.} 
\end{figure}

@@

Selection in the variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$. Density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in the variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

@@

We now turn to the issue of how density responds to selection, comparing the variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}. As we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery. To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.

@@

By contrast, density trajectories for $c$-selection in the variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density. As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent.

@@

The previous section underscores the fact that density must be affected by ongoing selection to threaten Eq.~\eqref{eq:canonical}. We now turn to selection on density-regulating traits such as $b$ and $d$ in the variable density lottery.

As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$. Thus, to threaten the assumption of density-independent $s$ in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.

This is the case for $b$ in the variable-density lottery. However, before we discuss this relatively complicated trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}. This only affects the sweep if $N$ changes substantially as the population shifts to the new equilibrium. From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occurs if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start and end of the sweep. (b) Example equilibrium-to-equilibrium sweep.}
\end{figure}

Equilibrium-to-equilibrium $b$-sweeps in the variable-density lottery are qualitatively different from $\delta$ sweeps, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps. Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.

If selection acts simultaneously on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in the variable-density lottery model as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
\end{figure}

@@

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.

>>>>

For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would then be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use the continuous time Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic growth rates. However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density. Thus, assuming viability, the increase in population density must be checked by crowding, implying that the Malthusian parameters $r_i$ eventually decline to zero (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

A rather different picture emerges in more ecologically explicit studies of selection in density-regulated populations. Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. Indeed, both $N$ and $s$ change during, and as a result of, adaptive sweeps in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).

Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is  pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

Nevertheless, studies of density-dependent selection rarely incorporate reproductive excess. This requires that we model a finite, density-dependent excess, which is substantially more complicated than modeling either zero (e.g. logistic) or infinite (e.g. Wright-Fisher) reproductive excess. Nei's ``competitive selection'' model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each genotype.

In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp. 54]{charlesworth_1994}. Reproductive excess is a special case corresponding to a critical pre-reproductive age group. A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp. 148]{charlesworth_1994}. This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case). By focusing on density in the critical age group, this literature loses the interdependence of pre-reproductive selection and post-reproductive density.

We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a finite reproductive excess. Our starting point is the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$. In each iteration of the classic lottery model, each type produces a large number of juveniles, such that $N$ remains constant (infinite reproductive excess). Competitive ability $c$ affects the probability of winning a territory, and behaves like a pure relative fitness trait. Thus, fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). Our first task is to relax the large-juvenile-number assumption of the lottery model to create a variable-density lottery with a finite, density-dependent reproductive excess.

The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above. The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density, and even when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that only one of the traits $b$, $c$, or $d$ is involved.

@@

We restrict our attention to asexual haploids, since it is then clearer how the properties of selection are tied to the underlying population ecological assumptions. We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce. All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$. We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Since we do not consider random genetic drift here, we will not analyze the fluctuations around these two expectations.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these propagules secure the territories without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, together with at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness). The third term, $A_i c_i/\overline{c}$, represents competitive victories in territories where two or more $i$ type propagules are present. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Two types are present with $c_1=1$, $c_2=1.5$ and $l_2/l_1=0.1$. Simulations are conducted as follows: $x_1,x_2$ values are sampled $U=10^5$ times from Poisson distributions with respective means $l_1,l_2$, and the victorious type in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/(c_1 x_1 + c_2 x_2)$. Dashed lines show the failure of the classic lottery model at low density.} 
\end{figure}

@@

Selection in our variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$. Density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

@@

We now turn to the issue of how density responds to selection, comparing our variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}. As we saw in the previous section, $c$-selection has no effect on population density in our variable-density lottery. To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.

@@

By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}. As a result, Eq.~\eqref{eq:canonical} holds in demographic equilibrium even though $c$-selection is density-dependent.

@@

For Eq.~\eqref{eq:canonical} to break down, the selection coefficient $s$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$. Eq.~\eqref{eq:canonical} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model. Thus, to threaten Eq.~\eqref{eq:canonical}, we require selection to be density-dependent, and also density to be changing. This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium. In this case the validity of Eq.~\eqref{eq:canonical} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonical} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
\end{equation}
Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep. (b) Example equilibrium-to-equilibrium sweep.}
\end{figure}

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection coefficients is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change. During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps. Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.

If selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}). This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{multiple.pdf}
\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in our variable-density lottery model as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
\end{figure}

@@

Summarizing the three traits in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly. This behavior is quite different from that found in the classical density-dependent selection literature (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection''). In the latter, selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients (``$\alpha$-selection''; \citealt{gill_1974,joshi_2001}). $K$ and $\alpha$ both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that strong selection is sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}).
